Hybrid Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-19 manuscript/main.tex
literature review updated

		\begin{abstract}
			Modern biomedical studies are increasingly collecting both functional and vector-valued predictors for each subject, introducing the new challenge of cross-modality correlations. To address this, we propose a hybrid partial least squares (PLS) regression framework that integrates functional and scalar predictors within a unified Hilbert space.
			We then extend the classical nonlinear iterative PLS (NIPALS) algorithm to this hybrid Hilbert space by iteratively maximizing the empirical cross-covariance between the hybrid predictor and the response. As a result, our method identifies low-dimensional representations that capture both within- and between-modality variation, as well as the response-predictor correlation. The procedure is computationally efficient, requiring only the solution of linear systems at each step. We provide theoretical properties to justify our algorithm and demonstrate its effectiveness through simulations and an application to clinical outcome prediction using renal imaging and scalar covariates from the Emory University renal study.
		\end{abstract}

@@

	%Infinite dimensional beta estimation problem (ill-posed problem)
	%Multicollinearity and high-dimensionality in scalar predictors are not handled.
	%Also, there may be high correlation between functional and scalar predictors.
	This hybrid model poses several challenges. 
	First, estimating the infinite-dimensional slope function $\beta_k(t)$ from a finite number of data
	is an ill-posed problem.% that requires dimension reduction or regularization based on structural assumptions, such as derivatives and covariance patterns.
	Second, we face the high-dimensionality of the predictors with $K+p$ dimensions.
	Third, strong correlations between functional and scalar predictors are common but often overlooked by separate modeling approaches.
	This paper unifies the remedies proposed for these three issues
by proposing a hybrid partial least square (PLS) regression framework that integrates functional and vector predictors in a principled and coherent manner. Before we summarize the key contributions of this new framework, we will first discuss previous approaches and their limitations.

@@

	For the high-dimensionality, a major solution is using derived inputs.
	A widespread approach in this line is the principal component analysis (PCA) regression.
	It is a two-step approach, applying (multivariate) functional PCA  to the functional predictors, possibly with standard PCA to the scalar predictors,
	followed by a classical multivariate regression on the resulting scores.
	PCA regression is well-studied for functional predictors alone \citep{hallMethodologyConvergenceRates2007, reissFunctionalPrincipalComponent2007, febrero-bandeFunctionalPrincipalComponent2017}.
	However, since the PCA-derived inputs are not informed by the response variable,   they are not guaranteed to capture the core regression relationship with a small number of derived inputs.
In terms of predictive power, partial least squares (PLS) regression is a powerful alternative. This two-step approach first iteratively constructs a set of orthogonal latent components from the predictors that have the maximal covariance with the response variable. This low-dimensional representation is then used to fit a linear regression model.
PLS for linear regression in the context of functional data was initiated by \citet{predaPLSRegressionStochastic2005} for signle predictor.
  PLS has been extensively studied , both for single and multiple functional predictors \citep{, reissFunctionalPrincipalComponent2007, aguileraUsingBasisExpansions2010, delaigleMethodologyTheoryPartial2012, aguileraPenalizedVersionsFunctional2016, febrero-bandeFunctionalPrincipalComponent2017, beyaztasRobustFunctionalPartial2022, saricamPartialLeastsquaresEstimation2022, mutisFunctiononfunctionLinearQuantile2025}.
	However, all of these existing PCA or PLS approaches overlook the potentially strong correlations between functional and scalar components. This can lead to multicollinearity and suboptimal predictive performance.
	To address the third challenge, \citet{jangPrincipalComponentAnalysis2021}  proposed a joint PCA method that accounts for correlations between functional and scalar data. However, the resulting components are still not informed by the response, so they may fail to capture the directions of maximal correlation with the outcome.

	To bridge this gap, we unify the remedies proposed for these three issues: basis expansion, partial least squares, and a hybrid approach that accounts for the correlation between functional and scalar parts.
	We propose a hybrid PLS regression framework that integrates functional and vector predictors in a principled and coherent manner. To extract predictive structure from these jointly observed and potentially correlated data types, we define a Hilbert space that treats the tuple of functional and vector components as a single hybrid object, equipped with a suitable inner product. The hybrid PLS direction is then obtained by iteratively maximizing the empirical covariance with the response, subject to a unit-norm constraint in this Hilbert space. Our framework is readily applicable to dense and irregular functional data and supports regularization techniques to prevent overfitting and reduce variance. We also provide the mathematical properties that justify our algorithm.

@@

	To extend Algorithm \ref{alg:scalar_pls} to the hybrid regression model \eqref{eq:hybrid_regression_model}, we must compute hybrid objects in both the PLS direction estimation and predictor residualization steps (lines \ref{alg:step:scalar_pls_direction} and \ref{alg:step:scalar_pls_residualization}). These computations cannot be separated into functional and scalar parts, as their correlation must be addressed.  
	A naive approach is to add loops over the observed evaluation points $t$ to lines \ref{alg:step:scalar_pls_direction} and \ref{alg:step:scalar_pls_residualization}, treating  
	\[
	\bigl( X_{i1}(t), \ldots, X_{iK}(t), Z_{i1}, \ldots, Z_{ip} \bigr) \in \mathbb{R}^{K+P}
	\]  
	as a Euclidean predictor, applying steps \ref{alg:step:scalar_pls_direction} and \ref{alg:step:scalar_pls_residualization}, and aggregating the results. This pointwise method, however, is computationally prohibitive for densely observed functional predictors, infeasible for irregular data, and prone to variability across the domain, resulting in overfitting and unstable predictions.  
	The main challenge of extension to the hybrid setting lies in the following points. First, 
	independent variables consist of multiple highly structured images and scalar predictors. 
	Second, our sample size is small compared to the dimension and number of functional and scalar predictors.
	Third,  existing partial least squares (PLS) methods can only accommodate (i) univariate or multivariate functional predictors without any scalar predictors \citep{predaPLSRegressionStochastic2005, delaigleMethodologyTheoryPartial2012, Febrero2017, Beyaztas2020}; or (ii) a univariate functional predictor with other scalar predictors \citep{Wang2018}.
	The next section proposes a new extension framework that addresses these issues.

@@

>>>>

		\begin{abstract}
	Regression of a scalar response on mixed functional- and scalar-valued predictors, such as medical imaging with auxiliary patient information, introduces the new challenge of handling cross-modality correlations. To address this, we propose a hybrid partial least squares (PLS) regression framework that integrates functional and scalar predictors within a unified Hilbert space.
			We then extend the classical nonlinear iterative PLS (NIPALS) algorithm to this hybrid Hilbert space by iteratively maximizing the empirical cross-covariance between the hybrid predictor and the response. As a result, our method identifies low-dimensional representations that capture both within- and between-modality variation, as well as the response-predictor correlation. The procedure is computationally efficient, requiring only the solution of linear systems at each step. We provide theoretical properties to justify our algorithm and demonstrate its effectiveness through simulations and an application to clinical outcome prediction using renal imaging and scalar covariates from the Emory University renal study.
		\end{abstract}

@@

	%Infinite dimensional beta estimation problem (ill-posed problem)
	%Multicollinearity and high-dimensionality in scalar predictors are not handled.
	%Also, there may be high correlation between functional and scalar predictors.
	This hybrid model poses several challenges. 
	First, estimating the infinite-dimensional slope function $\beta_k(t)$ from a finite number of data
	is an ill-posed problem.% that requires dimension reduction or regularization based on structural assumptions, such as derivatives and covariance patterns.
	Second, we face the high-dimensionality of the predictors with $K+p$ dimensions.
	Third, strong correlations between functional and scalar predictors are common but often overlooked by separate modeling approaches.
This paper addresses these three issues in a unified way by introducing a hybrid partial least squares (PLS) regression framework defined on a novel hybrid Hilbert space. Before summarizing the key contributions of this framework, we first review previous approaches and their limitations.

@@

	For the high-dimensionality, a major solution is using derived inputs.
A straightforward approach in this direction is principal component analysis (PCA) regression, which applies PCA separately to the functional- (by, for example, \citealt{happ_multivariate_2018}) and scalar predictors, and then performs a classical multivariate regression on the combined scores.
	PCA regression is well-studied for linear regression with functional predictors alone \citep{hallMethodologyConvergenceRates2007, reissFunctionalPrincipalComponent2007, febrero-bandeFunctionalPrincipalComponent2017}.
	However, since the PCA-derived inputs are not informed by the response variable,   they are not guaranteed to capture the core regression relationship with a small number of derived inputs.
In terms of predictive power, partial least squares (PLS) regression is a powerful alternative. It iteratively constructs a set of orthogonal latent components from the predictors that have the maximal covariance with the response variable, and use the resulting scores for regression.
%
PLS for linear regression in the context of functional data was introduced by \citet{predaPLSRegressionStochastic2005} for the case of a single predictor, studied from a stochastic process perspective. Motivated by regression on chemometric spectra, \citet{reissFunctionalPrincipalComponent2007}, \citet{aguileraUsingBasisExpansions2010}, and \citet{aguileraPenalizedVersionsFunctional2016} extended the framework by incorporating basis approximations and roughness penalties to promote smoothness. \citet{delaigleMethodologyTheoryPartial2012} provided the first thorough theoretical analysis, while \citet{saricamPartialLeastsquaresEstimation2022} proposed a computationally efficient procedure based on Golub–Kahan bidiagonalization. For a comprehensive review, see \citet{febrero-bandeFunctionalPrincipalComponent2017}. More recently, \citet{beyaztasRobustFunctionalPartial2022} extended the framework to accommodate multiple functional predictors.

However,  these existing approaches overlook the potentially strong correlations between functional and scalar components, which can lead to multicollinearity and suboptimal predictive performance. Multimodal correlations have mostly been addressed within an unsupervised learning framework. For instance, \citet{kolar_graph_2014} studied the estimation of joint undirected graphical models for  functional and vector data, while \citet{gengJointNonparametricPrecision2020} considered joint precision matrix estimation for brain measurements and confounding scalar covariate. \citet{jangPrincipalComponentAnalysis2021} proposed a joint PCA method that accounts for correlations between functional and scalar data. However, the resulting components are still not informed by the response and may fail to capture  correlation with the outcome.

	\subsection{Our contributions}
To address the gaps mentioned above, we unify the remedies proposed for these three issues: basis expansion, partial least squares, and accounting for correlations between the functional and scalar components.
	We propose a hybrid PLS regression framework that integrates functional and vector predictors in a principled and coherent manner. To extract predictive structure from these jointly observed and potentially correlated data types, we define a Hilbert space that treats the tuple of functional and vector components as a single hybrid object, equipped with a suitable inner product. The hybrid PLS direction is then obtained by iteratively maximizing the empirical covariance with the response, subject to a unit-norm constraint in this Hilbert space. Our framework is readily applicable to dense and irregular functional data and supports regularization techniques to prevent overfitting and reduce variance. We also provide the mathematical properties that justify our algorithm.

@@

	To extend Algorithm \ref{alg:scalar_pls} to the hybrid regression model \eqref{eq:hybrid_regression_model}, we must compute hybrid objects in both the PLS direction estimation and predictor residualization steps (lines \ref{alg:step:scalar_pls_direction} and \ref{alg:step:scalar_pls_residualization}). These computations cannot be separated into functional and scalar parts, as their correlation must be addressed.  
	A naive approach is to add loops over the observed evaluation points $t$ to lines \ref{alg:step:scalar_pls_direction} and \ref{alg:step:scalar_pls_residualization}, treating  
	\[
	\bigl( X_{i1}(t), \ldots, X_{iK}(t), Z_{i1}, \ldots, Z_{ip} \bigr) \in \mathbb{R}^{K+P}
	\]  
	as a Euclidean predictor, applying steps \ref{alg:step:scalar_pls_direction} and \ref{alg:step:scalar_pls_residualization}, and aggregating the results. This pointwise method, however, is computationally prohibitive for densely observed functional predictors, infeasible for irregular data, and prone to variability across the domain, resulting in overfitting and unstable predictions.  
	The main challenge of extension to the hybrid setting lies in the following points. First, 
	independent variables consist of multiple highly structured images and scalar predictors. 
	Second, our sample size is small compared to the dimension and number of functional and scalar predictors.
	Third,  existing partial least squares (PLS) methods can only accommodate (i) univariate or multivariate functional predictors without any scalar predictors \citep{predaPLSRegressionStochastic2005, delaigleMethodologyTheoryPartial2012, febrero-bandeFunctionalPrincipalComponent2017, Beyaztas2020}; or (ii) a univariate functional predictor with other scalar predictors \citep{Wang2018}.
	The next section proposes a new extension framework that addresses these issues.

@@

  \begin{theorem}[$L^2$ Convergence of Hybrid PLS] \label{thm:hybrid_pls_convergence} Let $Y_i$ be a scalar response and $W_i$ be a hybrid predictor in the hybrid Hilbert space $\mathbb{H} = (\mathbb{L}^2[0,1])^K \times \mathbb{R}^p$. Let $\hat{Y}_i = \langle \beta, W_i \rangle_{\mathbb{H}}$ denote the ordinary linear regression solution, where $\beta \in \mathbb{H}$ is the true regression coefficient. Let $\hat{Y}_{i,L} = \sum_{l=1}^L \hat{c}^{[l]} \rho_i^{[l]}$ be the Hybrid PLS approximation with $L$ components, where $\rho_i^{[l]} = \langle W_i, \xi^{[l]} \rangle_{\mathbb{H}}$ is the $l$-th PLS score and $\xi^{[l]} \in \mathbb{H}$ is the $l$-th PLS direction. Then, the Hybrid PLS approximation converges to the ordinary linear regression solution in the mean-squared ($L^2$) sense: $$ \lim_{L \to \infty} E\left[ \| \hat{Y}_{i,L} - \hat{Y}_i \|_{L^2(\Omega)}^2 \right] = 0 $$ \end{theorem} \begin{proof} The proof proceeds in three main steps: first, establishing that the PLS directions form a complete orthonormal system (CONS) for the relevant predictor subspace; second, expressing the ordinary linear regression solution as an infinite series in this PLS basis; and third, showing that the mean-squared error of the PLS approximation vanishes as the number of components increases. \subsection*{1. The PLS Directions Form a Complete Orthonormal System (CONS)} The Hybrid PLS algorithm iteratively constructs a set of directions $\{\xi^{[l]}\}_{l=1}^\infty$ from the predictors. The $l$-th PLS direction, $\xi^{[l]}$, is obtained as the principal eigenfunction of the operator $\mathcal{U}^{[l]} = \Sigma_{YW}^{[l]} \otimes \Sigma_{YW}^{[l]}$, where the superscript $[l]$ denotes that the operator is constructed from the residualized response $Y^{[l]}$ and predictors $W^{[l]}$ from the previous step. \paragraph{1.1 Orthonormality through Residualization} The iterative nature of the PLS algorithm guarantees the orthonormality of the directions. The \textbf{residualization (or orthogonalization)} step is crucial. After the $(l-1)$-th PLS direction $\xi^{[l-1]}$ is determined, the predictors and response are adjusted to remove the linear effect of this component. Specifically, for sample $i$: $$W_i^{[l]} = W_i^{[l-1]} - \langle W_i^{[l-1]}, \xi^{[l-1]} \rangle_{\mathbb{H}} \xi^{[l-1]}$$ $$Y_i^{[l]} = Y_i^{[l-1]} - \hat{c}^{[l-1]} \rho_i^{[l-1]}$$ where $W_i^{[1]} = W_i$ and $Y_i^{[1]} = Y_i$. This process ensures that the residualized predictors $W_i^{[l]}$ are orthogonal to all previous directions $\xi^{[j]}$ for $j < l$. Consequently, the newly found direction $\xi^{[l]}$, being derived from $W_i^{[l]}$, will be orthogonal to $\{\xi^{[1]}, \ldots, \xi^{[l-1]}\}$. By convention, PLS directions are normalized to have unit norm ($\Vert \xi^{[l]} \Vert_{\mathbb{H}} = 1$). Thus, the set $\{\xi^{[l]}\}_{l=1}^\infty$ is an orthonormal set. \paragraph{1.2 Completeness} The PLS algorithm, by repeatedly extracting the direction of maximum covariance from the residual space, exhaustively captures all linear information within the predictors that is correlated with the response. This iterative process terminates when the residualized predictors no longer exhibit covariance with the residualized response. Therefore, the span of the PLS directions, $\text{span}\{\xi^{[l]}\}_{l=1}^\infty$, precisely constitutes the subspace of $\mathbb{H}$ that contains all the linear information in the predictors relevant to predicting the response. This implies that $\{\xi^{[l]}\}_{l=1}^\infty$ forms a CONS for this relevant predictor subspace. \subsection*{2. The Ordinary Linear Regression Solution as a Series Expansion in the PLS Basis} Since $\{\xi^{[l]}\}_{l=1}^\infty$ is a CONS for the relevant predictor subspace, the true hybrid regression coefficient $\beta \in \mathbb{H}$ can be expressed as an infinite series in this basis: $$ \beta = \sum_{l=1}^{\infty} \langle \beta, \xi^{[l]} \rangle_{\mathbb{H}} \xi^{[l]} $$ Substituting this expansion into the definition of the ordinary linear regression solution $\hat{Y}_i$: $$ \hat{Y}_i = \langle \beta, W_i \rangle_{\mathbb{H}} = \left\langle \sum_{l=1}^{\infty} \langle \beta, \xi^{[l]} \rangle_{\mathbb{H}} \xi^{[l]}, W_i \right\rangle_{\mathbb{H}} $$ By the linearity of the inner product in a Hilbert space: $$ \hat{Y}_i = \sum_{l=1}^{\infty} \langle \beta, \xi^{[l]} \rangle_{\mathbb{H}} \langle \xi^{[l]}, W_i \rangle_{\mathbb{H}} $$ Let $c^{[l]} = \langle \beta, \xi^{[l]} \rangle_{\mathbb{H}}$ be the true coefficient for the $l$-th component and $\rho_i^{[l]} = \langle W_i, \xi^{[l]} \rangle_{\mathbb{H}}$ be the $l$-th PLS score for sample $i$. Then, $$ \hat{Y}_i = \sum_{l=1}^{\infty} c^{[l]} \rho_i^{[l]} $$ This shows that the exact regression solution is an infinite series involving the PLS scores. The PLS approximation $\hat{Y}_{i,L}$ is a partial sum of this series. Furthermore, it can be shown that the estimated coefficients $\hat{c}^{[l]}$ in the PLS approximation $\hat{Y}_{i,L}$ are consistent estimators of the true coefficients $c^{[l]}$. \subsection*{3. Proof of $L^2$ Convergence} Now, we examine the mean-squared error between the PLS approximation and the ordinary linear regression solution: $$ E\left[ \| \hat{Y}_{i,L} - \hat{Y}_i \|_{L^2(\Omega)}^2 \right] = E\left[ \left\| \sum_{l=1}^{L} c^{[l]} \rho_i^{[l]} - \sum_{l=1}^{\infty} c^{[l]} \rho_i^{[l]} \right\|^2 \right] $$ This can be rewritten as the expectation of the squared norm of the "tail" of the series: $$ E\left[ \| \hat{Y}_{i,L} - \hat{Y}_i \|_{L^2(\Omega)}^2 \right] = E\left[ \left\| - \sum_{l=L+1}^{\infty} c^{[l]} \rho_i^{[l]} \right\|^2 \right] $$ Since the PLS scores $\{\rho_i^{[l]}\}_{l=1}^\infty$ form an orthogonal system in the space of random variables (due to the orthogonality of $\{\xi^{[l]}\}$ and properties of expectation), the expectation of the squared sum simplifies to the sum of the expectations of the squared individual terms: $$ E\left[ \| \hat{Y}_{i,L} - \hat{Y}_i \|_{L^2(\Omega)}^2 \right] = \sum_{l=L+1}^{\infty} (c^{[l]})^2 E[(\rho_i^{[l]})^2] $$ The infinite series $\sum_{l=1}^{\infty} (c^{[l]})^2 E[(\rho_i^{[l]})^2]$ represents the total variance explained by the predictors and is a convergent series (related to the total variance of $\hat{Y}_i$). For any convergent series, the sum of its tail (i.e., from $L+1$ to infinity) must approach zero as $L$ tends to infinity. $$ \lim_{L \to \infty} \sum_{l=L+1}^{\infty} (c^{[l]})^2 E[(\rho_i^{[l]})^2] = 0 $$ Therefore, $$ \lim_{L \to \infty} E\left[ \| \hat{Y}_{i,L} - \hat{Y}_i \|_{L^2(\Omega)}^2 \right] = 0 $$ This concludes the proof that the Hybrid PLS approximation converges to the ordinary linear regression solution in the mean-squared sense. \end{proof}
